Gompertz growth as infinitely correlated non-local growth

### 2024-04-18 arxiv2/new_draft.tex
address Theas comments

\begin{abstract}
Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood. 
To add to this biological foundation, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. 
The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth rate decreases linearly, the trademark of Gompertz growth. We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction. 
This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities. 
This field could be interpreted as an environmental stressor like air-pollution, radiation, etc, or an environmental stimulant like an electromagnetic field. 
Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

\section*{Highlights}
\begin{itemize}
\item Logistic and Gompertz growth is connected with a novel framework based on microscopic principles.
\item Gompertz growth is connected with ubiquitous growth-activating field justified by physical and mathematical principles.
\item Growth processes is put in a larger contextual framework based on varying dependence in spacetime, including the Exponential, Gaussian and Poisson distributions. 
\end{itemize}

@@

In 1825, Gompertz published a paper on the nature of mortality where he discovered that from adulthood and onward ``the power to avoid destruction'' decreases exponentially with age. Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}. Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time \cite{makeham1889further}. Gompertz' law has later been known as the Gompertz model or Gompertz growth.

@@

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Some biological macroscopic growth phenomena are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}.

In fact, it has been shown that cancer malignancies are mostly \textit{negative} in electric polarity with respect to the rest of the body, while nonmalignant (or benign) tumor pathology is mostly \textit{positive} in polarity \cite{langman1949technique}. Gompertz growth has also been widely observed in cancerous cells \cite{kozusko2003combining,laird1964dynamics}. One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth are coupled with an electromagnetic field of specific character. Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.

On the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to be excluded from a subset of models in the communicable disease paradigm. They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range with the communicable disease assumption does not cover the parameter value taken in the Gompertz scenario. If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near-Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}. 
We will show that one can obtain a true Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
% Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.

\subsection{Main Results and Roadmap}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. In particular, we posit that the source of growth can be elucidated by considering interactions in the microscopic domain which can be aggregated to recover the traditional macroscopic growth equations. Second, we will elaborate on the idea that the logistic model results from local interactions while the Gompertz model results from a field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz and Poisson.

Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which offers an alternative to the traditional Richards model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)),
\end{equation}
where $\beta>0$ is a growth parameter. Here, the right hand side is commonly interpreted as representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \frac{\beta}{\theta} (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards model \citep{richards1959flexible}. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximally cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

In spite of the connection of the Gompertz model with higher order terms in an augmented logistic model, it is not clear whether the source of growth is external or internal to the system. The right hand side ``resource'' term is not specific enough so as to specify the location of the source. One could argue that the since the Gompertz model has asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in Markov's language) that initiates the growth and subsequently decays exponentially like a memoryless process. To further explore the source of growth we turn to the microscopic domain.

@@

Now as we let N grow, we impose a simple rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. This condition translates into an adjacency matrix with each column containing a single non-zero entry. 
Using this rule and rearranging and summing across all entities we obtain the following relationship,
\begin{equation}
\label{eq:modSIR22}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

@@

Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on. Analogous to the pairwise exclusivity rule, each entity can again only participate in \textit{one} interaction group at any given order and any given direction. For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level (Figure \ref{fig:order}).

@@

Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously. 
In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism. Examples of this field could be sound waves or the electromagnetic field.

Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms is coming from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case. 
To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same logistic dependency can be obtained under the exclusivity condition. This could be interpreted as a causality reversal of the growth process.
Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model (see Appendices \ref{appendix:swappinglogistic} and \ref{appendix:swappinggompertz}).

Causality reversal is indeed what is done when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g. \citet{estrada2022networked}). But because higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head. In this context, the Gompertz model would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. Thus, with the necessary requirement of higher order interactions between nodes, \textit{the Gompertz model cannot be used as a communicable disease model}.
% While some applied work show how scale-free pairwise interactions approximate Gompertz, theoretical parity is not shown (e.g. \citet{zonta2022virus}).

\section{Unifying the Gaussian and Exponential curves with the Logistic and the Gompertz curves}

@@

Solving this equation yields the familiar Gaussian function, 
\begin{equation}
  s(t) = A e^{ - \beta t^2 / 2},
\end{equation}
where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$. Thus, not surprisingly Gaussian growth emerges for mutually independent events that have a linear time dependence. The same result was originally formulated by Gauss with space as the independent variable instead of time and with spatial estimation error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details). In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.

If we remove the dependency on time altogether, we observe the exponential growth equation
\begin{equation}
  \frac{d}{dt}\ln s(t) = - \beta,
\end{equation}
which yields 
\begin{equation}
  s(t) = A e^{ - \beta t}.
\end{equation}
This exponential equation implies a complete independence on both time and of the mutual entities involved in the growth. In the context of measurement error this function would be the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}. 
% Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.

@@

@@

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection where the Poisson distribution reflects a coherent system. Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here. The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.

@@

An infinitely interacting system where entities are classical in nature with well-defined position and momentum, breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}. With a such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli. Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019). \citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic in nature, and while Lauro et al. do not discuss the external nature of such a source, their model does not preclude it. More generally, with growth as stochastically rooted it is also ubiquitously present uniformly across all entities from the very onset of growth. Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.

@@

We finally note that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above would be the living conditions on our planet, for which time acts as a proxy. One could further postulate this stressor as a field surrounding the earth, e.g. in its electromagnetic footprint often related with the Schumann Resonance \cite{cherry2002schumann,schumann1952strahlungslosen}. It is indeed curious that we do not see mortality rates behave as linearly dependent on time, for which a Gaussian curve would be appropriate. Nor do we see any finite-ordered polynomial time dependence on mortality rates. Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality coming from an environmental stressor.

>>>>

\begin{abstract}
Logistic and Gompertz growth has traditionally been connected by augmenting the logistic model with an extra parameter to obtain what is known as $\theta$-logistic growth or the Richards model for growth. In spite of this bridge, the biological foundation of Gompertz growth is only vaguely understood. 
To add to this biological foundation, we show that there is another way to connect the two growth modes by adding higher order terms in the growth equation, and present both a macroscopic and microscopic representation of growth where the latter involves a network of entities. 
The microscopic representation elucidates the source of Gompertz growth as the potential abundance and not the realized abundance, and an example of such a network is presented. 
For finite higher order terms we show that the growth equation can be represented by Gauss' Hypergeometric function, and only for infinitely many terms does the abundance lend itself conveniently to a log-transformation wherein the log-abundance growth rate decreases linearly, the trademark of Gompertz growth. We thus equate Gompertz growth with correlated microscopic entities that exhibit long-range or non-local interaction. 
This conclusion adds to the interpretation of biological systems that undergo Gompertz growth where long-range interaction could be interpreted as the source of growth being external to the individual entities, like a field that stimulates growth simultaneously across all entities. 
This field could be interpreted as an environmental stressor like air-pollution, electromagnetic radiation, etc.
Our new discovery allows for a unification of well-known growth models as follows: time independent and uncorrelated growth yields the exponential growth function, time dependent and uncorrelated growth yields Gaussian growth, time independent and pairwise correlated growth yields the logistic function and time independent and infinitely correlated growth yields the Gompertz function.

@@

\section*{Highlights}
\begin{itemize}
\item Logistic and Gompertz growth is connected within a novel framework based on microscopic principles.
\item Gompertz growth is connected with a ubiquitous growth-activating field justified by physical and mathematical principles.
\item Growth processes are put in a larger contextual framework based on varying dependence in spacetime, including the Exponential, Gaussian and Poisson distributions. 
\end{itemize}

@@

In 1825, Gompertz published a paper on the nature of mortality where he discovered that from adulthood and onward ``the power to avoid destruction'' decreases exponentially with age. Equivalently, mortality rates increase in geometric progression for arithmetic increases in age, an observation now commonly referred to as Gompertz' law of mortality \cite{Gompertz1825}. Makeham (1889) later provided an interpretation to Gompertz' law where the recuperative force of a human, or ``vital force'', becomes less efficient with time as a ``stressor'' \cite{makeham1889further}. Gompertz' law has later been known as the Gompertz model or Gompertz growth.

@@

It is worth noting that the notion that Gompertz growth is maximally coherent and non-local has also been derived from quantum mechanical principles \cite{molski2003coherent}. Molski and Konarski (2003) showed that the Gompertz equation is the solution of a form of the time-independent Schrodinger equation. From this result, they derive the quantum coherent states which are non-local in space while moving along a classical time trajectory. This adds to our exhibits that the Gompertz model can be interpreted as non-local in origin. Coherent quantum states was originally formulated in the context of electromagnetic fields by \citet{glauber1963coherent} (for which he received the Nobel Prize). Some biological macroscopic growth phenomena are also due to such fields. For example, the role of internal electrical fields in regenerative growth is well established and is based on negative electrical potentials that act on sensitive cells \cite{becker1984electromagnetic}. In fact, it has been shown that cancer malignancies are mostly \textit{negative} in electric polarity with respect to the rest of the body, while nonmalignant (or benign) tumor pathology is mostly \textit{positive} in polarity \cite{langman1949technique}. Gompertz growth has also been widely observed in cancerous cells \cite{kozusko2003combining,laird1964dynamics}. One is then led to the interpretation that certain biological growth processes that exhibit Gompertz growth possibly are coupled with an electromagnetic field of specific character. Consequently and unsurprisingly, the artificial disruption of such a field has been found to disrupt the mitotic cell cycle \cite{goodman1983pulsing,liboff1981alternating,norton1984bioelectric,liboff1984time}.

On the topic of disease modelling, the Gompertz model has been shown by Wang et al. \cite{wang2012richards} to be incompatible with a subset of models in the communicable disease paradigm. They used the common Susceptible-Infected-Recovered (SIR) model by \citet{kermack1927contribution} augmented by the Richards parameter and showed that the allowed parameter range within the this model does not cover the parameter value taken in the Gompertz scenario. If one incorporates a social proximity network on which the disease spreads, \citet{zonta2022virus} observed through computer simulations that a near-Gompertz growth can occur but only for scale-free networks, a type of network that has been shown to be rare in human social networks \cite{clauset2009power,broido2019scale,holme2019rare}. 
We will show that one can obtain a true Gompertz curve from a ubiquitous stressor scenario based on microscopic interactions.
% Given these observations, any mortality pattern that shows a Gompertz characteristic seems not be a result of a communicable disease, but rather from a ubiquitous stressor. 
While many studies show the SIR model (and its extensions) as the fitted model in epidemics \cite{carletti2020covid,cooper2020sir,postnikov2020estimation,munoz2021sir,cooper2022dynamical,saikia2021covid}, some recent studies have successfully fitted a Gompertz model \cite{Ohnishi2020,Rypdal2020,Catala2020,rodrigues2020monte,Levitt2020}.

\subsection{Main aim}
In this paper, our primary aim is to investigate the source of growth in self-organized processes. In particular, we posit that the source of growth can be elucidated by considering interactions in the microscopic domain which can be aggregated to recover the traditional macroscopic growth equations. Second, we will elaborate on the idea that the logistic model results from local internal interactions while the Gompertz model results from a possibly external field or equivalently from an infinite set of higher order interactions between the microscopic entities involved in the growth. We will devise a slightly different parametrization than the Richards model, the generalized Gompertz-Logistic equation which is commonly used to compare the two models (e.g. by \citet{petroni2020logistic}, \citet{tjorve2017use}, and \citet{wang2012richards}). In particular, instead of considering a decimal number as the unifying exponent we will appeal to integer exponents with a number of terms that tend towards infinity as we approach the Gompertz model. As we shall see, this is rooted in the microscopic interpretation of the growth system, where each subgroup of $N$ elementary entities interacts with all other equally sized subgroups each composed of mutually exclusive elementary entities. If these subgroups are present at all orders of cardinality, the Gompertz model emerges, while only pairwise interactions yield the logistic model. As seen in the aforementioned references, this is in line both with experimental and theoretical results. In presenting this argument, we present a novel bridge between the logistic and the Gompertz model which offers both a mathematical and biological foundation to the source of growth. Lastly, we will present a unifying framework in the growth context for the common functions Gaussian, Exponential, Logistic, Gompertz, and Poisson by considering independence between entities and other spatio-temporal dependencies on growth.

\subsection{Roadmap}
Our paper is structured as follows: we will first present a new macroscopic foundation of the Gompertz-Logistic spectrum which offers an alternative to the traditional Richards model. Then, we present a microscopic foundation which acts as the building blocks of the macroscopic phenomena. This is where we show that the Gompertz model results from an infinite set of higher order interactions between the elementary entities involved in the growth, which could alternatively be seen as a ubiquitous field or in the biological context: an environmental stressor. We proceed with a discussion that summarizes our findings in the context of the other well-known growth modalities involving the Gaussian and the Laplacian model and then conclude.

@@

First consider a system that grows according to logistic growth. If we let the variable $X(t)$ denote the abundance of a quantity in a normalized system where its maximum size is 1, the equation that determines its size as a function of time is
\begin{equation}
  \frac{1}{X}\frac{d}{dt} X(t) = \beta (1 - X(t)),
\end{equation}
where $\beta>0$ is a growth parameter. Here, the right hand side is commonly interpreted as representing ``resources available for growth'' which decrease linearly with respect to the relative abundance's growth rate on the left hand side. One way to generalize logistic growth is to introduce a parameter that exponentiates the resource availability in the following way,
\begin{equation}
\label{eq:Rich}
  \frac{1}{X}\frac{d}{dt} X(t) = \frac{\beta}{\theta} (1 - X^{\theta}(t)),
\end{equation}
also called $\theta$-logistic growth or Richards model \citep{richards1959flexible}. Petroni et al. (2020) argued that this exponentiation can be interpreted as non-linear interaction effects where there is a level of cooperation in the microscopic domain. They showed that in the limit where the system becomes \textit{maximally cooperative} or \textit{coherent}, $\theta\rightarrow 0$ and the $\theta$-logistic equation reduces in this limit to the Gompertz growth model,
 \begin{equation}
 \label{eq:Gomp}
  \frac{1}{X}\frac{d}{dt} X(t) = - \beta \ln X(t).
\end{equation}
In this limit, the relative growth rate goes to infinity as $\lim_{t\to 0}$, which could be seen as a verification of the maximally coherent characteristic of the system. One is reminded of a system that is perturbed and left to reach a new equilibrium. Petroni's definition of maximal coherence in the context of the Gompertz equation was shown to rigorously correspond to the definition of coherence used in quantum mechanical systems \cite{molski2003coherent} attributed to Glauber (1963) \cite{glauber1963coherent}.

@@

In spite of the connection of the Gompertz model with higher order terms in an augmented logistic model, it is not clear whether the source of growth is external or internal to the system. The right hand side ``resource'' term is not specific enough so as to specify the location of the source. One could argue that since the Gompertz model has an asymptotically infinite initial growth rate, the system could be seen as exposed to a ubiquitous signal (or catalyst in the language of \citet{markov2019reaction}) that initiates the growth and subsequently decays exponentially like a memoryless process \citep{durrett1999essentials}. To further explore the source of growth we turn to the microscopic domain.

@@

Now as we let N grow, we impose an exclusivity rule: \textit{Each entity can only receive growth potential from a single entity and create growth potential in a single possibly different entity}. This condition translates into an adjacency matrix with each column containing a single non-zero entry. 
Using this rule and rearranging and summing across all entities we obtain the following relationship,
\begin{equation}
\label{eq:modSIR22}
\frac{d}{dt} \ln \left[ \prod_i^N x_i \right ] = \sum_j^N s_{j}.
\end{equation}

@@

Instead of limiting the growth equation to pairwise interactions, we allow higher order interactions in the form of triplets, quadruplets, and so on. Analogous to the pairwise exclusivity rule, \textit{each entity can again only participate in one interaction group at any given order and any given direction}. For example, any triplet can only interact with one other triplet and have exclusive possession of its entities at the triplet level (Figure \ref{fig:order}).

@@

Since all higher order interaction would necessarily imply non-local interaction, the natural interpretation would be that the growth is activated by a ubiquitous field accessible by all entities simultaneously. 
In this way one could think of the activating field as an \textit{effectively} external field, even though it is possible that the field emerges as a collective emergent phenomenon purely from within the system or organism. Examples of such a field could be sound waves or an electromagnetic field.

Another way to see the source of growth as effectively external is to recognize that the source of growth in the higher order terms originate from the abundance potential and not the realized potential, while in the logistic growth model this is not necessarily the case. 
To see this, first swap the role of $x_i$ and $s_i$ on the right hand side of Equation \ref{eq:modSIR} to see that the same logistic dependency can be obtained under the exclusivity condition. This could be interpreted as a causality reversal of the growth process.
Then note that when augmenting the growth equation with higher order terms on the right-hand side of the same equation, the same swapping will not lead to the Gompertz model (see Appendices \ref{appendix:swappinglogistic} and \ref{appendix:swappinggompertz}).

Causality reversal is indeed necessary when modelling communicable diseases in a microscopic network context, where $x_i$ represents the infected nodes and $s_i$ represents the susceptible nodes (see e.g. \citet{estrada2022networked}). Higher order growth does not allow this causality reversal, where the infected nodes are the source of spread, the Gompertz model cannot be used since it would imply that the susceptible nodes are the source of the disease spreading, which would turn communicable disease theory on its head. In this context, the emergence of Gompertz growth would rather suggest a system disturbed by a ubiquitous, simultaneous, and non-local stressor eliciting a corresponding stress response through which a new stress tolerance baseline is gradually established. Thus, with the necessary requirement of higher order interactions between nodes, \textit{the Gompertz model cannot be used as a communicable disease model}.
% While some applied work show how scale-free pairwise interactions approximate Gompertz, theoretical parity is not shown (e.g. \citet{zonta2022virus}).

\section{Unifying the Gaussian, Exponential, and Poisson models with the Logistic and the Gompertz models}

\subsection{Unification with the Gaussian and Exponential models}

@@

Solving this equation yields the familiar Gaussian function, 
\begin{equation}
  s(t) = A e^{ - \beta t^2 / 2},
\end{equation}
where $\beta$ can be interpreted as the inverse variance or dispersion of the growth phenomena and where $A=1$ when $s(0)=1$. Thus, not surprisingly, Gaussian growth emerges for mutually independent events that have a linear time dependence. The same result was originally formulated by Gauss with space as the independent variable instead of time and with spatial estimation error as the dependent variable instead of abundance, all in the context of the prediction of the heavenly object called Ceres \cite{gauss1857theory} (see \cite{stahl2006evolution} for details). In the context of measurement error, one would equate the above equation to the statement that the relative measurement error decreases linearly with size.

@@

If we remove the dependency on time altogether, we observe the exponential growth equation
\begin{equation}
  \frac{d}{dt}\ln s(t) = - \beta,
\end{equation}
which yields 
\begin{equation}
  s(t) = A e^{ - \beta t}.
\end{equation}
This exponential equation implies a complete independence on both time and of the entities involved in the growth. In the context of measurement error this function would represent the Laplacian error which states that absolute measurement error decreases linearly with the error itself \footnote{Or as Laplace originally stated: ``[...] we have no reason to suppose a different law for the ordinates than for their differences'' \cite{stahl2006evolution}.}. 
% Note also that for both the Gaussian and the exponential growth functions, the \textit{realized abundance}, $x(t)$, could be substituted as the dependent variable where the physical interpretation would be a decay process.

@@

As a side note, the Poisson distribution can be included in this discussion as the distribution which governs the infinitesimal growth rate of a coherent or infinitely correlated group of entities. By letting the growth rate be the rate at which a zero-count is observed with the Poisson mean parameter equal to $-\beta t$, one recovers the Gompertz model in Equation \ref{eq:Gomp}. This argument was put forth by \citet{shklovskii2005simple}, and \citet{glauber1963coherent} was the first to point out the connection of the Poisson distribution with a coherent system. Glauber also showed that incoherent systems, or systems with independent entities, follow the Gaussian distribution which is in accordance with our results and discussion here. The mapping between coherence and incoherence has also been postulated as a medical diagnostic technique by \citet{zhang1994log}, where incoherence is a sign of disease while coherence is a sign of well-being.

@@

Growth of infinitely interacting systems where entities are classical in nature with well-defined position and momentum breaks the law of causality unless we allow the source of growth to be a ubiquitous field, externally or internally created, which affects all microscopic entities instantaneously \footnote{Or at speeds much greater than the differential times in the growth equations}. With such a source, entities need not be interacting with each other but rather respond to the field-induced stimuli. Examples of such models are already given by De Lauro (2014), Sacher (1956) and Markov (2019). \citet{de2014stochastic} and \citet{sacher1956statistical} consider the source of growth to be stochastic or random in nature, and while Lauro et al. do not discuss the external nature of such a source, their model does not preclude it. More generally, with growth as stochastically rooted the source is also ubiquitously present uniformly across all entities from the very onset of growth. Markov (2019)\cite{markov2019reaction} discusses growth within the context of chemical reactions, where the source of growth is a catalytic agent that is implicitly available ubiquitously to all reactants to the same degree, exactly in line with De Lauro and Sacher's models.

@@

We finally note that the original formulation of the Gompertz model made by Gompertz himself was concerning mortality rates as a function of age. He noticed that mortality rates increased at a geometric rate with age. In this context the ubiquitous stressor postulated above would be the living conditions on our planet, for which time acts as a proxy. It is indeed curious that we do not see mortality rates behave as linearly dependent on time, for which a Gaussian curve would be appropriate. Nor do we see any finite-ordered polynomial time dependence on mortality rates. Instead we see growth of mortality rates as sourced from other entities at all interaction orders, which in light of our discussion above is equivalent to mortality triggered by an environmental stressor.
